Viscous Erosion in Porous Media

### 2018-04-03 docs/methods/erosion_methods.tex
NM

	Next, we aim to measure the opening angle of the front/rear corner. Since these simulated bodies are actually smooth on a fine scale, we must have a procedure to estimate the apparent angle that results from the high (but finite) curvature at the stagnation points. To this end, we fit the tangent angle $\theta(\alpha)$ with a 7th-degree polynomial over a region excluding the stagnation points ($\alpha = 0$ and 0.5). We then extrapolate the polynomial to the stagnation points, which gives an estimate of the angle there. Although this procedure is subject to some uncertainty (to be examined below), we have found it to give robust angle measurements. Figure \ref{fig:arangle}(b) shows the opening angle, $\beta$, as it varies in time for the three cases of smoothing parameters. For the initially circular geometry, our procedure yields the correct opening angle of $180^{\circ}$, i.e.~a flat front face. As time proceeds, erosion carves a sharper corner and $\beta$ decreases monotonically. Here, the three simulation curves are more easily distinguished, implying that the opening angle is more sensitive to the level of numerical smoothing than is the aspect ratio. The curves nevertheless show convergence as $\eps, \sigma \to 0$. In the finest simulation ($\eps = \sigma = 5/1024$), $\beta$ converges to the value $102^{\circ}$, which agrees precisely with the analytical prediction from Section \ref{LimitingShape}.

	To more thoroughly examine the influence of the smoothing parameters, we report in Table~\ref{table:arangle} the long-time values of the aspect ratio and opening angle for each of the three simulations. The table shows that both the aspect ratio and opening angle converge with first-order as $\eps, \sigma \to 0$. In addition, we show an estimate of the uncertainty involved in our angle measurement, obtained by varying the degree of the fitting polynomial and the size of the excluded region around the stagnation points. The estimates show that as $\eps, \sigma \to 0$ the angle-measurements become more precise and converge to the predicted value of $102^{\circ}$.

@@

	In summary, we have found that a single-body eroding in 2D Stokes flow converges to a terminal form characterized by nearly uniform shear stress along its surface. The final shape is fairly slender, with an aspect ratio of 2.7, and distinguished by  corners of opening angle $102^{\circ}$ at the front and rear. This shape is closely related to the drag-minimizing profile found by Pironneau (1973) --- the geometry that minimizes drag in 3D Stokes flow subject to the constraint of fixed volume \cite{pir1973}. Using variational principles, Pironneau showed that the drag-minimizing form has constant shear (or vorticity) along its surface. The resulting {\em axisymmetric} shape has an aspect ratio of 2.1 and front/rear opening-angles of $120^{\circ}$ \cite{pir1973, mit-spa2016}. Interestingly, 3D simulations of viscous erosion showed close resemblance between the shape formed by a single eroding body and this Pironneau form \cite{mit-spa2016}. It was found that the eroding body does not converge {\em precisely} to the Pironneau shape because the long-time surface-stress distribution created by erosion is only {\em approximately} uniform. Rather than strict uniformity, variations in body curvature must be offset by slight variations in $\Vn$ for self-similar reduction \cite{pir1973, mit-spa2016}. Pironneau's argument that the drag-minimizing form has constant surface stress applies to 2D as well. Therefore, the shape observed to develop in our simulations is approximately the form that minimizes drag subject to fixed area, i.e.~the 2D counterpart to Pironneau's shape. How close this shape is to the true drag-minimizer is related to how close the stress distributions are to uniform, as can be seen in Fig.~\ref{shrink_intface}(b).

@@

We now discuss some physical aspects of multi-body erosion revealed by our simulations. Figure~\ref{fig:06bodies} shows a simulation of 6 eroding bodies, each initially circles of various sizes and positions. Over time, erosion not only diminishes the size of the bodies, but also alters their shapes considerable. As in the single-body simulations, corner-like features develop, but not in locations that are easily predictable. Rather, when and where these sharp features develop depends on inter-body feedback as mediated by the Stokes flow.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.9 \textwidth]{./figs/06bod.pdf}
\caption{\label{fig:06bodies} 6 bodies.
}
%A time lapse of 50 bodies eroding in a viscous flow.  The boundary conditions are a constant pressure drop that creates a flow from the left to right.  The rate of erosion is proportional to the shear stress which is equivalent to the vorticity evaluated on the bodies.  The colors represent the vorticity, so regions with large vorticity (in magnitude) are regions where the flow will erode the fastest.
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure generated from data file 50circ512p5, 
%which uses 50circ512.in with epsfac = 30, sigfac = 10, dt = 1e-4, fixpdrop = 1

@@

The initial configuration in  Fig.~\ref{fig:06bodies} was selected to have several near contacts between bodies, for example between bodies 1 \& 2, 3 \& 5, and 5 \& 6. Our method handles these near contacts without difficulty. Interestingly, erosion appears to preferentially remove material from these regions of near contact, as the narrow spacing between bodies expands rather rapidly. For example, the spacing between bodies 1 \& 2 is initially very small, but by the second frame, is comparable to the spacing between bodies 3 \& 4. Even more interesting, the bodies tend to develop flat faces where there are near contacts. This effect leads to the appearance of straight channels in between bodies, for example between bodies 1 \& 2, 3 \& 5, and 5 \& 6. Once formed, these straight channels persist until the bodies have nearly vanished.

The vorticity field not only allows us to visualize regions of high shear, but also allows us to identify preferred channels with most intense flow. Figure~\ref{fig:06bodies} shows that the flow is concentrated primarily in a single channel that runs horizontally below bodies 1 \& 2 and above bodies 3, 5, \& 6. This main channel widens the fastest, and secondary channels follow (e.g.~between bodies 1 \& 2 or 3 \& 5 or 5 \& 6). Since the pressure drop across the cell is fixed, the overall flow rate increases as the bodies erode. This can be seen by the overall vorticity magnitude growing in time. By the final frame shown, body 4 has vanished completely and our method handles this event seamlessly.

>>>>

	Next, we aim to measure the opening angle of the front/rear corner. Since these simulated bodies are actually smooth on a fine scale, we must have a procedure to estimate the apparent angle that results from the high (but finite) curvature at the stagnation points. To this end, we fit the tangent angle $\theta(\alpha)$ with a 7th-degree polynomial over a region excluding the stagnation points ($\alpha = 0$ and 0.5). We then extrapolate the polynomial to the stagnation points, which gives an estimate of the angle there. Although this procedure is subject to some uncertainty (to be examined below), we have found it to give robust angle measurements. Figure \ref{fig:arangle}(b) shows the opening angle, $\beta$, as it varies in time for the three cases of smoothing parameters. For the initially circular geometry, our procedure yields the correct opening angle of $180^{\circ}$, i.e.~a flat front face. As time proceeds, erosion carves a sharper corner and $\beta$ decreases monotonically. Here, the three cases of smoothing parameters are more easily distinguished, implying that the opening angle is more sensitive to the level of numerical smoothing than is the aspect ratio. The curves nevertheless show convergence as $\eps, \sigma \to 0$. In the finest simulation ($\eps = \sigma = 5/1024$), $\beta$ converges to the value $102^{\circ}$, which agrees precisely with the analytical prediction from Section \ref{LimitingShape}.

	To more thoroughly examine the influence of the smoothing parameters, we report in Table~\ref{table:arangle} the long-time values of the aspect ratio and opening angle for each of the three cases. The table shows that both the aspect ratio and opening angle converge with roughly first-order as $\eps, \sigma \to 0$. In addition, we show an estimate of the uncertainty involved in our angle measurement, obtained by varying the degree of the fitting polynomial and the size of the excluded region around the stagnation points. The estimates show that as $\eps, \sigma \to 0$ the angle-measurements become more precise and converge to the predicted value of $102^{\circ}$.

@@

	In summary, we have found that a single-body eroding in 2D Stokes flow converges to a terminal form characterized by nearly uniform shear stress along its surface. The final shape is fairly slender, with an aspect ratio of 2.7, and distinguished by  corners of opening angle $102^{\circ}$ at the front and rear. This shape is closely related to the drag-minimizing profile found by Pironneau (1973) --- the geometry that minimizes drag in 3D Stokes flow subject to the constraint of fixed volume \cite{pir1973}. Using variational principles, Pironneau showed that the drag-minimizing form has constant shear (or vorticity) along its surface. The resulting {\em axisymmetric} shape has an aspect ratio of 2.1 and front/rear opening-angles of $120^{\circ}$ \cite{pir1973, mit-spa2016}. Interestingly, 3D simulations of viscous erosion showed close resemblance between the shape formed by a single eroding body and this Pironneau form \cite{mit-spa2016}. It was found that the eroding body does not converge {\em precisely} to the Pironneau shape because the long-time surface-stress distribution created by erosion is only {\em approximately} uniform. Rather than strict uniformity, variations in body curvature must be offset by slight variations in $\Vn$ for self-similar reduction \cite{pir1973, mit-spa2016}. Pironneau's argument that the drag-minimizing form has constant surface stress applies to 2D as well. Therefore, the shape observed in our simulation is approximately the form that minimizes drag subject to fixed area, i.e.~the 2D counterpart to Pironneau's shape. How close this shape is to the true drag-minimizer is related to how close the stress distributions are to uniform, as can be seen in Fig.~\ref{shrink_intface}(b).

@@

With the validation complete, we now discuss some of the emergent physics of multi-body erosion revealed by our simulations. We begin with a relatively simple case of 6 eroding bodies, as this will allow us to identify key principles at work in more complex cases (e.g.~the 50-body simulation in Fig.~\ref{fig:50bodies}). Figure~\ref{fig:06bodies} shows the simulated erosion of 6 bodies, each initial circles of various sizes and positions. Over time, erosion not only diminishes the size of the bodies, but also alters their shapes considerable. As in the single-body simulations, corner-like features develop, but not in locations that are easily predictable. Rather, when and where these corners form depends on how neighboring bodies interact as mediated by the Stokes flow.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.9 \textwidth]{./figs/06bod.pdf}
\caption{\label{fig:06bodies} The simulated erosion of 6 bodies with the fixed-pressure drop condition. Erosion causes flat faces to develop wherever bodies were initially very close to one another. This effect leads to the appearance of straight channels between bodies, and causes the bodies themselves to appear distinctly polygonal. The vorticity field (color-bar at right) shows that the flow is concentrated primarily in a single channel that runs horizontally across the domain. }
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure generated from data file 50circ512p5, 
%which uses 50circ512.in with epsfac = 30, sigfac = 10, dt = 1e-4, fixpdrop = 1

The initial configuration in  Fig.~\ref{fig:06bodies} was selected to have several points at which the bodies nearly contact one another, for example bodies 1 \& 2, 3 \& 5, and 5 ur \& 6. Intriguingly, erosion appears to remove material preferentially from these near-contact regions, as the narrow spacing between neighboring bodies expands rather rapidly. For example, the spacing between bodies 1 \& 2 is initially very small, but by the second frame, is comparable to the spacing between bodies 3 \& 4. Furthermore, the bodies tend to develop flat faces at the locations of near contact, leading to the appearance of straight channels between bodies. Straight channels can be seen, for example, between bodies 1 \& 2, 3 \& 5, and 5 \& 6. Once formed, these channels persist until the bodies have nearly vanished. Notice that by the final frame, body 4 has vanished completely, and the simulation handles this event without difficulty.

The vorticity field helps not only to visualize regions of high shear, but also to identify the certain preferred channels in which the flow is most intense. As seen in Fig.~\ref{fig:06bodies}, the flow is concentrated primarily in a single channel running horizontally below bodies 1 \& 2 and above bodies 3, 5, \& 6. This main channel widens the fastest, and secondary channels (e.g.~between bodies 1 \& 2 or 3 \& 5 or 5 \& 6) follow. Notice that, in these simulations, the overall vorticity magnitude increases in time as a result of fixing the pressure-drop across the cell.

@@
